End-User Customization by Direct Manipulation of Tabular Data

### 2020-05-23 paper.tex
tweak
more pictures

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Customizing Software by Direct Manipulation of Tabular Data}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  In this paper we show how the behavior of a software application can
  be extended and adapted by direct manipulation. We introduce
  \emph{table-driven customization}, a new paradigm where end users
  directly manipulate a table view of the structured data inside an
  application, rather than writing imperative scripts (as in most
  customization tools). This simple model also accommodates a
  spreadsheet formula language and custom data editing widgets, which
  provide sufficient expressivity to implement many useful
  customizations.

  We illustrate the approach with Wildcard, a browser extension that
  implements table-driven customization in the context of existing web
  applications. Through concrete examples, we show that this paradigm
  can support useful customizations for real websites. We share
  reflections from our experiences using the Wildcard system, on both
  its strengths and limitations relative to other customization
  approaches. Finally, we explore how this paradigm might lead to new
  software architectures that encourage this form of customization.
\end{abstract}

@@

Many applications don't meet the precise needs of their users. End user
customization systems can help improve the situation, by empowering
non-programmers to modify their software in the way they would like.

Most end user customization systems offer a simplified version of
programming. Some scripting languages \citep{bolin2005, cook2007} have a
friendly syntax that resembles natural language. Other visual
customization tools eliminate text syntax entirely. Macro recorders
\citep{cook2007, chasins2018, anupam2000} remove some of the initial
programming burden by letting a user start with concrete demonstrations.
Despite their many differences, these approaches all share something in
common: an imperative programming model, with statement sequencing,
mutable variables and loops.

We have known for decades about an alternate approach: \emph{direct
manipulation} \citep{shneiderman1983}, where ``visibility of the object
of interest'' replaces ``complex command language syntax''. Direct
manipulation is the \emph{de facto} standard in GUIs today, but when it
comes to customizing those GUIs, it is rarely to be found. As a result,
switching from using software to customizing it poses a learning barrier
for users not familiar with programming, and requires an abrupt shift in
mental model even for those familiar with scripting.

In this work, we ask: what would it look like to build a direct
maniulation interface that lowers the threshold for starting to
customize software? We take inspiration from spreadsheets and visual
database query interfaces, which have successfully enabled millions of
end users to compute with data through direct manipulation.

In this paper we present a technique called \emph{table-driven
customization} inspired by those tools. An application's UI is augmented
with a table view, where the user can see and manipulate the
application's internal data. Changes in the table view result in
immediate corresponding changes to the original user interface, enabling
the user to customize an application with live feedback.

@@

We have developed a browser extension called Wildcard which uses web
scraping techniques to implement table-driven customization for existing
Web applications. In Section~\ref{sec:example}, we illustrate the end
user experience of table-driven customization through an example
scenario using Wildcard.

@@

We have successfully used Wildcard to build customizations for 11
different websites. In Section~\ref{sec:reflections}, we present
reflections from this process, outlining the kinds of customizations we
were able to build, limitations we encountered, and the challenges of
integrating scraping logic with websites.

@@

To illustrate the end user experience of table-driven customization, we
consider an example scenario of customizing Hacker News, a tech news
aggregator. Figure~\ref{fig:hacker-news} shows accompanying screenshots.

@@

The table contains additional columns to the right of the original
attributes for each row, where the user can enter spreadsheet-style
formulas to compute derived values. We provide generic names for these
columns by default as spreadsheets do, so that users don't need to
choose column names themselves. The user enters a formula into the
\texttt{user1} column: (Figure~\ref{fig:hacker-news}, Note C):

@@

This calls a built-in function that takes a URL as an argument and uses
a public web API to compute an estimated read time for the link's
contents. The argument \texttt{link} refers to a column name in the
table; for each row, the formula is evaluated with the link for that
particular row. The result of evaluating this formula is a numerical
value in seconds for each article.

@@

@@

\emph{Returning a table}: A table adapter exposes a table of data: an
ordered list of records. Each record carries a unique identifier and
associates named attributes with values. Tables have a typed schema, so
the same attributes are shared across all records. The columns also
carry some additional metadata in addition to their type, such as
whether or not they are read-only or editable.

@@

\hypertarget{types-of-table-adapters}{%
\subsubsection{Types of table adapters}\label{types-of-table-adapters}}

We have built three specific types of table adapters so far, to enable
web customization using Wildcard.

\textbf{DOM scraping adapters} are the essential component that enables
Wildcard to interface with an existing website UI. In addition to the
standard web scraping problem of extracting a table of data from the
DOM, a scraping adapter must also manipulate the DOM to reorder rows,
edit form entries, and inject annotations as the table is edited.

@@

After specifying basic metadata like which URLs the adapter should apply
to, the programmer only needs to implement a single Javascript function
that scrapes relevant elements from the page. They are free to use any
scraping techniques; in practice simple CSS selectors and DOM APIs
usually suffice. The generic adapter then wraps this scraping function
to implement the table adapter interface. For example, when the table
view is sorted, the generic adapter takes the DOM elements corresponding
to the rows in the table, removes all of them from the page, and then
reinserts them in the new order.

The programmer can optionally override other default behavior of the
generic DOM scraping adapter, including the logic for when to re-run the
scraping function in response to page changes, how to style injected
annotations, and how to style the currently selected row in the table.

An \textbf{AJAX scraping adapter} intercepts AJAX requests made by a web
page, and extracts information from those requests to add to the table.
When available, this tends to be a helpful technique because the data is
already in a structured form so it is easier to scrape, and it often
includes valuable information not shown in the UI.

@@

The \textbf{local storage adapter} simply stores a table of data in the
browser. This is currently only used to persist annotations: the user
makes edits in the table, they are stored privately in the browser, and
can be loaded into the table on future pageloads.

@@

\begin{table*}[]
\begin{tabular}{llrl}
\hline
\textbf{Website} & \textbf{Description} & \textbf{LOC} & \textbf{Example customizations}                                                              \\ \hline
Airbnb           & Travel               & 73                                       & Add Walkability Scores to listings. Sort listings by price.                           \\
Amazon           & Online shopping      & 99                                       & Sort used book sellers by total price, including delivery fees                                \\
Blogger          & Blogging             & 36                                       & Use alternate text editor to edit blog posts                                                 \\
Expedia          & Travel               & 41                                       & Use alternate datepicker to enter travel dates                                               \\
Flux             & Data portal          & 67                                       & Use Wildcard as a faster table editor for editing lab results                                \\
Github           & Code repository      & 62                                       & Sort a user's code repositories by stars to find popular work                                \\
Hacker News      & News                 & 69                                       & Add read times to links. Filter out links that have been read. \\
Instacart        & Grocery delivery     & 48                                       & Sort groceries by price and category. Take notes on items.                                   \\
Uber Eats        & Food delivery        & 117                                      & Sort/filter restaurants by estimated delivery ETA and price.                                 \\
Weather.com  & Weather              & 51                                       & Sort/filter hourly weather to find nice times of day.                                        \\
Youtube          & Videos               & 80                                       & Sort/filter videos by length, to find short videos to watch.                                 \\ \hline
\end{tabular}
\caption{A list of table-driven customizations that we have implemented using Wildcard.}
\label{tab:websites}
\end{table*}

@@

We built Wildcard integrations for 11 websites, including transactional
sites like Amazon and Uber Eats, and media consumption sites like Hacker
News and Youtube. Table 1 summarizes these integrations, showing the
number of lines of code in the adapter configuration for each site, and
some example customizations we were able to perform for each site.

@@

@@

Many web annotation systems focus on annotating text or arbitrary
webpage content, but Wildcard limits annotations to structured objects
extracted by an adapter, resulting in a different set of use cases.
Annotating with Wildcard has proven most useful when taking notes on a
list of possible options (e.g., evaluating possible Airbnb locations to
rent). We have also used it with Instacart's online grocery cart, for
jotting down notes like ``should we get more milk?''

@@

Formulas are especially useful for fetching data from Web APIs. We've
used them to augment Airbnb listings with walkability scores, and to
augment Hacker News articles with estimated read times. One challenge of
the current language design is that supporting a new web API requires
writing Javascript code to add a new function to the language, because
web APIs typically return complex JSON data structures that can't be
easily displayed in a single table cell. In the future we would like to
make it possible to call new APIs without adding a dedicated function,
which might require adding functions to the formula language that can
manipulate JSON data.

@@

@@

AJAX scraping proved highly useful in several cases. The Uber Eats
website was challenging to scrape because it has a complex DOM structure
with machine-generated CSS classes, but the site also generates AJAX
requests which contain all the relevant data in a structured form that
is much easier to extract. We also found examples where relevant
information wasn't present in the DOM at all. On the grocery delivery
site Instacart, we used AJAX scraping to augment grocery items with
their categorization in the store, enabling use cases like viewing all
the vegetables in an order together.

@@

\emph{other platforms?}

>>>>

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{End-User Customization by Direct Manipulation of Tabular Data}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  In this paper we introduce \emph{table-driven customization}, a new
  way for end users to customize software. We augment a user interface
  with a table view showing the structured data inside the application.
  When users edit the table, their changes are reflected in the original
  UI. This simple model also accommodates a spreadsheet formula language
  and custom data editing widgets, providing enough power to implement a
  variety of useful extensions.

  We illustrate the approach with Wildcard, a browser extension that
  implements table-driven customization for existing applications using
  web scraping. Through concrete examples, we show that this paradigm
  can support extensions to many real websites, ranging from sorting and
  filtering existing data to adding entire new features. We share
  reflections from our experiences using Wildcard, on both its strengths
  and limitations relative to other customization approaches. Finally,
  we explore how this paradigm might lead to new software architectures
  that encourage this form of end-user customization.
\end{abstract}

@@

Many applications don't meet the precise needs of their users, and it is
impossible for developers to anticipate everyone's unique requirements.
End user customization systems can help close this gap, by empowering
non-programmers to modify their software to satisfy their personal
goals.

Many end user customization systems
\citep{cook2007, bolin2005, leshed2008, chasins2018} offer a scripting
model. They use various strategies to make programming more
approachable: friendly syntax, a visual programming environment, or
macro recording to bootstrap from concrete demonstrations. But all these
techniques build on the same fundamental foundation: an imperative
programming model, with statement sequencing, mutable variables, and
loops.

We have known for decades about an alternative: \emph{direct
manipulation} \citep{shneiderman1983}, where ``visibility of the object
of interest'' replaces ``complex command language syntax''. Direct
manipulation is the \emph{de facto} standard in GUIs today, but when it
comes to customizing those GUIs, it is rarely to be found. Switching
from using an interface to customizing it via scripting requires an
abrupt shift in interaction model, and can pose a steep learning barrier
for users not familiar with programming. We subscribe to MacLean et al's
vision of a ``gentle slope'' \citep{maclean1990}, where users should
only need to make minimal and incremental investments in skill to
achieve their desired customizations.

We seek to add a point to this gentle slope by proposing a new method
for customizing software via direct manipulation, which requires less
skill than scripting but still supports powerful customizations. We take
inspiration from visual database query interfaces and spreadsheets,
which have successfully enabled millions of end users to compute with
data through direct manipulation.

In this paper, we present a paradigm called \emph{table-driven
customization}. An application's UI is augmented with a table view where
the user can see and manipulate the application's internal data, just as
in a familiar spreadsheet. These changes don't just apply to the table;
they also result in immediate changes to the application's original user
interface. The user can sort/filter data in the UI, inject annotations
into the UI, pull in related information from other web services, and
more, all using the table as a mediating interface.

@@

We have developed a browser extension called Wildcard which uses web
scraping techniques to implement table-driven customization for existing
Web applications. In Section~\ref{sec:example}, we illustrate the end
user experience of table-driven customization through an example
scenario, using Wildcard to add features to Hacker News.

@@

We have successfully used Wildcard to build customizations for 11
different websites which serve our own personal needs. In
Section~\ref{sec:reflections}, we present reflections from this process.
We outline the kinds of customizations we were able to build,
limitations we encountered, and some of the challenges of writing
scraping logic.

@@

To illustrate the end user experience of table-driven customization, we
consider an example scenario of customizing
\href{https://news.ycombinator.com/}{Hacker News}, a tech news
aggregator. Figure~\ref{fig:hacker-news} shows accompanying screenshots.

@@

The table contains additional empty columns (\texttt{user1},
\texttt{user2}, \ldots), where the user can enter spreadsheet-like
formulas to compute derived values. The user enters a formula into the
first column: (Figure~\ref{fig:hacker-news}, Note C):

@@

This calls a built-in function that takes a URL as an argument and uses
a public web API to compute an estimated read time for the link's
contents. The \texttt{link} argument refers to a column name in the
table; the formula is automatically evaluated across all rows in the
table, using the value of \texttt{link} for each row.

@@

Here we describe the abstract interface fulfilled by a table adapter,
and then describe specific types of table adapters we've built so far in
the Wildcard system.

\hypertarget{abstract-interface}{%
\subsubsection{Abstract interface}\label{abstract-interface}}

@@

\emph{Returning a table}: A table adapter exposes a table of data: an
ordered list of records. Each record carries a unique identifier and
associates named attributes with values. Tables have a typed schema, so
the same attributes are shared across all records. We currently support
strings, numeric values, booleans, and datetimes as types. The columns
also carry some additional metadata, such as whether or not they are
read-only or editable.

@@

\hypertarget{dom-scraping-adapters}{%
\subsubsection{DOM scraping adapters}\label{dom-scraping-adapters}}

DOM scraping adapters are the essential component that enables Wildcard
to interface with an existing website UI. In addition to the standard
web scraping problem of extracting a table of data from the DOM, a
scraping adapter must also manipulate the DOM to reorder rows, edit form
entries, and inject annotations as the table is edited.

@@

\emph{Specifying scraping logic}: First, the adapter author specifies
basic metadata like which URLs the adapter should apply to, and a list
of columns and their types. Then, implement a single Javascript function
that extracts relevant elements from the page using web scraping
technqiues. The generic adapter then wraps this scraping function to
implement the table adapter interface. For example, when the user sorts
the table view, the generic adapter removes the DOM elements
corresponding to the rows in the table, and reinserts them back into the
page in the correct order.

For each scraped row of data, the programmer must return a unique
identifier, and a value for each column in the table.

\emph{Choosing an ID}: When possible, we recommend choosing a
server-side identifier that will remain stable across pageloads. This
enables user annotations persisted in local storage to be associated
with the same records on subsequent pageloads. We have found that it's
usually possible to find such an identifier; for example, each item in a
page often contains a link to a page with more details, where the URL
contains a unique identifier.

\emph{Types of scraped values}: For each individual value within a row,
there are two options for what type of data can be returned by the
programmer-specified scraping function.

The default option is to return a DOM element, in which case the generic
adapter extracts the text contents of the DOM element and casts them to
the type of the column. For example, if the relevant column is a numeric
column and the returned value is a \texttt{\textless{}div\textgreater{}}
containing the string ``1.23'', the generic adapter extracts the number
1.23. The advantage of this approach is that the field becomes
\emph{writable} in the table, if the adapter author chooses to mark it
as writable. When the user edits the value in the table, the generic
adapter can overwrite the inner contents of the DOM element to reflect
the change. In practice, this is most useful for editing the value of
form input fields. It's possible to enable the editing of other fields
(e.g., editing the title of a post on Hacker News) but we haven't found
many use cases for editing those types of fields.

Another option is to directly return a value, rather than returning a
DOM element. The advantage of this approach is that the adapter author
can perform arbitrary computations to derive the returned value---for
example, they can use a regular expression to extract a substring. The
disadvantage is that the field is no longer writable, because the
computation used to derive the value isn't reversible.

\emph{Other site-specific customization}: The programmer can optionally
override other behavior of the generic DOM scraping adapter, including
logic for when to re-run the scraping function in response to page
changes, how to style injected annotations, and how to style the
currently selected row in the table. We've chosen sensible defaults that
generally result in a workable adapter without needing to change these
options, but they can improve the experience on a specific website. For
example, by default we highlight the currently selected row with a blue
border, but on some sites this is hard to see or breaks the layout, so
setting a background color works better.

\hypertarget{ajax-scraping-adapters}{%
\subsubsection{AJAX scraping adapters}\label{ajax-scraping-adapters}}

An AJAX scraping adapter intercepts AJAX requests made by a web page,
and extracts information from those requests to add to the table. When
available, this tends to be a helpful technique because the data is
already in a structured form so it is easier to scrape, and it often
includes valuable information not shown in the UI.

@@

\hypertarget{local-storage-adapters}{%
\subsubsection{Local storage adapters}\label{local-storage-adapters}}

The local storage adapter simply stores a table of data in the browser.
This is currently only used to persist annotations.

The table view is initialized with empty columns such as \texttt{user1}
which serve as the user's ``scratch space,'' as shown in
Section~\ref{sec:example}. When the user makes edits to these columns,
new rows are created in the local storage table. The rows contain the
record ID from the DOM scraping adapter, which enables them to be
re-associated with the same records on subsequent pageloads.

@@

\begin{table*}[]
\begin{tabular}{llrl}
\hline
\textbf{Website} & \textbf{Description} & \textbf{LOC} & \textbf{Example customizations}                                                              \\ \hline
Airbnb           & Travel               & 73                                       & Add Walkability Scores to listings. Sort listings by price.                           \\
Amazon           & Online shopping      & 99                                       & Sort third party sellers by total price, including fees.                                \\
Blogger          & Blogging             & 36                                       & Use alternate text editor to edit blog posts.                                                 \\
Expedia          & Travel               & 41                                       & Use alternate datepicker to enter travel dates.                                               \\
Flux             & Data portal          & 67                                       & Use Wildcard as a faster table editor for editing lab results.                                \\
Github           & Code repository      & 62                                       & Sort a user's code repositories by stars to find popular work.                                \\
Hacker News      & News                 & 69                                       & Add read times to links. Filter out links that have been read. \\
Instacart        & Grocery delivery     & 48                                       & Sort groceries by price and category. Take notes on items.                                   \\
Uber Eats        & Food delivery        & 117                                      & Sort/filter restaurants by estimated delivery ETA and price.                                 \\
Weather.com  & Weather              & 51                                       & Sort/filter hourly weather to find nice times of day.                                        \\
Youtube          & Videos               & 80                                       & Sort/filter videos by length, to find short videos to watch.                                 \\ \hline
\end{tabular}
\caption{A list of table-driven customizations that we have implemented using Wildcard.}
\label{tab:websites}
\end{table*}

@@

We built Wildcard integrations for 11 websites, including transactional
sites like Amazon and Uber Eats, and media consumption sites like
Youtube and Hacker News. Table 1 summarizes these integrations, showing
the number of lines of code in the adapter configuration for each site,
and some example customizations we were able to perform for each site.

@@

\begin{figure}
\hypertarget{fig:amazon}{%
\centering
\includegraphics[width=\columnwidth]{media/amazon.png}
\caption{Sorting the used sellers page on Amazon by total price}\label{fig:amazon}
}
\end{figure}

\begin{figure}
\hypertarget{fig:ubereats}{%
\centering
\includegraphics[width=\columnwidth]{media/ubereats.png}
\caption{Organizing takeout restaurants on Uber Eats by ETA and price}\label{fig:ubereats}
}
\end{figure}

@@

Many web annotation systems focus on annotating text or arbitrary
webpage content, but Wildcard limits annotations to structured objects
extracted by an adapter, resulting in a different set of use cases.
Annotating with Wildcard has proven most useful when taking notes on a
list of possible options (e.g., evaluating possible Airbnb locations to
rent). We have also used it with Instacart's online grocery cart, for
jotting down notes like ``do we need more eggs?''

@@

Formulas are especially useful for fetching data from Web APIs. We've
used them to augment Airbnb listings with walkability scores, and to
augment Hacker News articles with estimated read times as shown in
Section~\ref{sec:example}. One challenge of the current language design
is that supporting a new web API requires writing Javascript code to add
a new function to the language, because web APIs typically return
complex JSON data structures that can't be easily displayed in a single
table cell. In the future we would like to make it possible to call new
APIs without adding a dedicated function, which might require adding
functions to the formula language that can manipulate JSON data.

@@

\begin{figure}
\hypertarget{fig:instacart}{%
\centering
\includegraphics[width=\columnwidth]{media/instacart.png}
\caption{Taking notes on Instacart grocery items, after sorting them by price}\label{fig:instacart}
}
\end{figure}

@@

AJAX scraping proved highly useful in several cases. The Uber Eats
website was challenging to scrape because it has a complex DOM structure
with machine-generated CSS classes, but the site also use AJAX requests
which contain all the relevant data in a structured form that is much
easier to extract. We also found examples where relevant information
wasn't present in the DOM at all. On the grocery delivery site
Instacart, we found that AJAX requests contained additional data not
shown in the UI, enabling us to do things like sort grocery items in an
order by category.

@@
